Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2021-12-24 analysis/paper/paper.Rmd
Minor edits to text

# Load custom plot functions
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())

#setOxcalExecutablePath("/tmp/RtmpFVOcBJ")
```

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g. @clark2017] and other dimensions that have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

# Some initial data preparation
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_nonflint)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_nonflint:flake_retouch_nonflint))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  tibble::rownames_to_column(var = "no")

@@

```{r tab1}

# Table. Overview of chronological framework.
Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
            'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9300–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
framework <- c(rep("Glørstad (2010)", 4), rep("Reitan (2016)", 5))

@@

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)) . This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance also means that the structure of the data is

@@

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}

@@

  # If the dating method is 14C, draw a sample using the associated posterior
  # pdf
  if(site["dating_method"][1,] == "c14"){
    samplingframe <- list(x = site[["date_grid"]], y = site[["probabilities"]])
    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
  # If not, draw a random year from the reported date range (uniform probability)
  } else {
    sdate <- sample(as.integer(site["reported_start_bc"]):as.integer(site["reported_end_bc"]), 1)
  }
  # Return sampled date with site name as rowname. 
  sdate <- data.frame(sdate)
  row.names(sdate) <- site[["site"]][1]
  return(sdate)
}

@@

# Supplementary material

```{r supp1}
# # Supplementary table of radiocarbon dates
# 
# select(radiocarbon, -excavation_project) %>% 
#   kable(booktabs = TRUE, longtable = TRUE,
#       col.names = c("Site name", "Material", "Lab code", "C14-age", "Error"),
#       linesep = c(""))
```

@@

# Reviewer 2 wanted non-flint as a separate column and the artefact categories
# combined, irrespective of raw-material. 
all_nonflint_prop <- sites %>%
    mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined + 
      projectile_nonflint,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch +
      knife_nonflint + scraper_nonflint + drill_nonflint + 
      microblade_retouch_nonflint + blade_retouch_nonflint + 
      fragment_retouch_nonflint +
      flake_retouch_nonflint,
    macro_tool = axe + chisel + axe_nonflint + 
      chisel_nonflint + club_hatchet_nonflint,
    core = core + core_nonflint,
    core_fragment = core_fragment + core_fragment_nonflint,
    axe_debitage = axe_debitage + axe_debitage_nonflint,
    blade = blade + blade_nonflint,
    microblade = microblade + microblade_nonflint,
    fragment = fragment + fragment_nonflint,
    flake = flake + flake_nonflint,
    chip = chip + chip_nonflint,
    nonflint_prop = nonflint_count / artefact_count,
    .keep = "unused") %>% 
    # Make site names the rownames
    column_to_rownames(var = "no") %>% 
    # Deselect a few of the columns not be used
    select(-excavation_project, -excavated_m3, -dating_method)

>>>>

# Load custom plot functions
source(here::here("analysis/script/plot_functions.R"), local = knitr::knit_global())
```

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g. @clark2017] and other factors that have been employed to align properties of lithic assemblages with land-use and mobility patterns.

@@

# Some initial preparation of the data
sites <- assemblages %>% 
  # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_nonflint)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_nonflint:flake_retouch_nonflint))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  tibble::rownames_to_column(var = "no")

@@

```{r tab1}
# Table. Overview of chronological framework.
Period <- c('Early Mesolithic, Fosna Phase', 'Middle Mesolithic, Tørkop Phase', 'Late Mesolithic, Nøstvet Phase', 'Late Mesolithic, Kjeøy Phase',
            'Flake Axe Phase', 'Microlith Phase', 'Pecked Adze Phase', 'Nøstvet Adze Phase', 'Transverse Arrowhead Phase')
Duration <- c("9500–8200 BCE", "8200–6300 BCE", "6300–4600 BCE", "4600–3800 BCE", "9300–8200 BCE", "8200–7000 BCE", "7000–5600 BCE", "5600–4500 BCE", "4500–3900 BCE")
framework <- c(rep("Glørstad (2010)", 4), rep("Reitan (2016)", 5))

@@

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). This for example pertains to tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. That these categories are internally consistent and categorically exclusive in terms of fulfilled purpose is at best a dubious proposition, in turn potentially rendering their contribution as discrete analytic units misleading. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots.

@@

```{r wabi, fig.cap = paste0("Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics (\\textit{r} = ", round(wabicor, 2), "), B) Flint (\\textit{r} = ", round(wabicorflint, 2), "). The logarithm is taken to base 10 on all axes."), fig.height = 5, fig.width = 10}

@@

  # If the dating method is 14C, draw a sample using the associated posterior
  # pdf
  if(site["dating_method"][1,] == "c14"){
    samplingframe <- list(x = site[["date_grid"]], y = site[["probabilities"]])
    sdate <- with(samplingframe, sample(x, size = 1, prob = y))
  # If not, draw a random year from the reported date range (uniform probability)
  } else {
    sdate <- sample(
      as.integer(
        site["reported_start_bc"]):as.integer(site["reported_end_bc"]), 1)
  }
  # Return sampled date with site name as rowname. 
  sdate <- data.frame(sdate)
  row.names(sdate) <- site[["site"]][1]
  return(sdate)
}

@@

Supplementary material

@@

# A reviewer wanted artefact types combined irrespective of raw-material and 
# non-flint as supplementary column. I start by including it as a normal 
# column.
all_nonflint_prop <- sites %>%
    mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined + 
      projectile_nonflint,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch +
      knife_nonflint + scraper_nonflint + drill_nonflint + 
      microblade_retouch_nonflint + blade_retouch_nonflint + 
      fragment_retouch_nonflint +
      flake_retouch_nonflint,
    macro_tool = axe + chisel + axe_nonflint + 
      chisel_nonflint + club_hatchet_nonflint,
    core = core + core_nonflint,
    core_fragment = core_fragment + core_fragment_nonflint,
    axe_debitage = axe_debitage + axe_debitage_nonflint,
    blade = blade + blade_nonflint,
    microblade = microblade + microblade_nonflint,
    fragment = fragment + fragment_nonflint,
    flake = flake + flake_nonflint,
    chip = chip + chip_nonflint,
    nonflint_prop = nonflint_count / artefact_count,
    .keep = "unused") %>% 
    # Make site names the rownames
    column_to_rownames(var = "no") %>% 
    # Deselect a few of the columns not be used
    select(-excavation_project, -excavated_m3, -dating_method)
